---
title: "exploring CLOVER"
author: "Daniel Suh"
date: "11/19/2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

CLOVER is an aggregate database of four different databases: GMPD, HP3, EID2, and Shaw et al.
https://github.com/viralemergence/clover
Let's see how it is to work with it.

Packages  
```{r, message = F}
library(tidyverse)
library(magrittr)
library(plotly)
library(here)
```

Read data  
```{r, message = F}
mammal_viruses <- read_csv(here("data/clover/CLOVER_0.1_MammalViruses_AssociationsFlatFile.csv"))
columns <- read_csv(here("data/clover/CLOVER_ColumnDescriptions.csv"))
```

Check it out  
```{r}
mammal_vsr <- mammal_viruses %>% select(Host, Virus) %>% distinct() %>% group_by(Host) %>% summarize(psr = n())
```


```{r}
gmpd <- read_csv(here("data/GMPD_main.csv"))
```

More data
```{r, message=F}
viruses <- read_csv(here("data/clover/CLOVER_1.0_Viruses_AssociationsFlatFile.csv"))
helminths <- read_csv(here("data/clover/CLOVER_1.0_HelminthProtozoaFungi_AssociationsFlatFile.csv"))
bacteria <- read_csv(here("data/clover/CLOVER_1.0_Bacteria_AssociationsFlatFile.csv"))
```

```{r}
colnames(mammal_viruses) <- gsub("Virus", "Pathogen", colnames(mammal_viruses))
mammal_viruses %<>% mutate(PathogenType = "virus")

pathogens <- rbind(viruses,bacteria,helminths,mammal_viruses)
pathogens <- distinct(pathogens)
summary(pathogens)
#table(pathogens$HostOrder)
#table(pathogens$PathogenType)
PSR <- pathogens %>% select(Host, Pathogen) %>% distinct() %>% group_by(Host) %>% summarize(psr = n())
path_others <- pathogens %>% filter(PathogenType=="others")
table(path_others$PathogenClass)
```




Bigyra is a phylum within the superphylum Heterokonta
Chromadorea are a class of roundworms (Nematoda)
Ichthyosporea are a class of eukaryotes that parasitize fish
Ichthyostracta are a class of crustaceans

Parasite Summary
```{r}
ggplot(pathogens, aes(PathogenType, fill = Database)) + geom_bar()
pathogens %>% filter(PathogenType=="bacteria/rickettsia") %>% ggplot(.,aes(PathogenClass, fill = Database)) +geom_bar()+theme(axis.text.x = element_text(angle=45,vjust=0.5))
pathogens %>% filter(PathogenType=="virus") %>% ggplot(.,aes(PathogenClass, fill = Database)) +geom_bar()+theme(axis.text.x = element_text(angle=45,vjust=0.6))
```

Host Summary
```{r}
ggplot(pathogens, aes(HostClass, fill = Database)) + geom_bar() + theme(axis.text.x = element_text(angle=70, vjust = 0.5))
pathogens %>% filter(HostClass=="mammalia") %>% ggplot(.,aes(HostOrder,fill = Database)) + geom_bar() + theme(axis.text.x = element_text(angle=70, vjust = 0.5))
ggplot(pathogens, aes(Database)) + geom_bar() + plotly()

pathogens %>% group_by(Database) %>% summarize(n=n()) %>% plot_ly(x= ~Database, y = ~n, type="bar") %>% layout(title="# Records by Database")
```

GMPD Comparison

Is the entire GMPD dataset within clover or does clover filter out some of the GMPD records or uses a less complete version?
```{r}
clover_gmpd <- pathogens %>% filter(Database=="GMPD2")
glover_PSR <- clover_gmpd %>% select(Host, Pathogen) %>% distinct() %>% group_by(Host) %>% summarize(psr = n())
gmpd_PSR <- gmpd %>% select(HostCorrectedName,ParasiteCorrectedName) %>% distinct() %>% group_by(HostCorrectedName) %>% summarize(psr=n())

ggplot(clover_gmpd, aes(PathogenType)) + geom_bar()
ggplot(gmpd, aes(ParType)) + geom_bar()
table(clover_gmpd$PathogenType)
table(gmpd$ParType)
```

looks like gmpd has more parasites than the clover version of gmpd. So the clover version of gmpd is not as comprehensive as actual gmpd but why??


# EID2

Are the records in EID2 in clover using the genbank data only or also uses the text mined data?
```{r}
eid_clover <- pathogens %>% filter(Database == "EID2")
eid_clover %>% ggplot(., aes(x=DetectionMethodOriginal)) + geom_histogram(stat = "count")
```

Hmm NCBI nucleotide seems like genbank while cargocarrier may just be the text mined ones.

```{r}
eid_clover %>% filter(DetectionMethodOriginal == "NCBI Nucleotide") %>% ggplot(., aes(x=PathogenType)) + geom_histogram(stat = "count")
eid_clover %>% filter(DetectionMethodOriginal == "NCBI Nucleotide") %>% ggplot(., aes(x=HostClass)) + geom_histogram(stat = "count")
eid_clover %>% filter(DetectionMethodOriginal == "NCBI Nucleotide") %>% filter(HostClass == "mammalia") %>% ggplot(., aes(x=HostOrder)) + geom_histogram(stat = "count") + theme(axis.text.x = element_text(angle = 70, hjust = 1))
```

How does clover's EID2 compare to the EID2 that I use?

```{r}
eid <- read_csv(here("previous work", "data", "source", "EID2_full.csv"))
```

```{r}
table(eid$`Cargo classification`)
table(eid_clover$PathogenType)

cleid_ncbi <- eid_clover %>% filter(DetectionMethodOriginal == "NCBI Nucleotide")
table(cleid_ncbi$PathogenType)
```


Hmmm EID2 from clover is wayyyy larger than the EID2 I have.




Check for sampling bias between datasets. How sensitive is PSR to number of records of database?




Parasite diversity matrix

Each row is a host

Each column is a parasite or parasite order

Counts are host-parasite associations by parasite order

Using phylogenetic data, an index of diversity can be measured for each host

```{r}
para_by_order <- pathogens %>% select(Host, HostOrder, Pathogen, PathogenOrder) %>% distinct() %>% group_by(Host, PathogenOrder) %>% summarize(n=n())
host_sp_para_order_mat <- para_by_order %>% pivot_wider(names_from = PathogenOrder, values_from = n)
```



