---
title: "GMPD Pantheria"
author: "Daniel Suh"
date: "`r Sys.Date()`"
output: html_document
---

# Important note

Functions from brt.functions.R require data to be in "data.frame" format. Make sure to reclass any data using as.data.frame() before running gbm.step



```{r packages}
library(tidyverse)
library(magrittr)
library(here)

library(caret)

source(here("previous work", "elith_tutorial", "brt.functions.R"))
```

Read in data. To start, we are focusing on gmpd and pantheria

```{r read}
gmpd <- read_csv(here("data/GMPD_main.csv"))
panth <- read_tsv(here("data/pantheria/ECOL_90_184/PanTHERIA_1-0_WR05_Aug2008.txt"))
```

We need to change these maxed out values (-999) to be NAs instead. We are also removing some variables if they are all NA or are not informative for our purposes (taxonomy)

```{r panth NAs}
panth[panth == -999.00] <- NA

colnames(panth) <- gsub("\\d+-\\d+_", "", colnames(panth))
colnames(panth) <- gsub("*\\d+", "", colnames(panth))


panth$ref_count <- str_count(panth$References, ";")

panth$`ActivityCycle` <- as.factor(panth$`ActivityCycle`)
panth$`DietBreadth` <- as.factor(panth$`DietBreadth`)
panth$`HabitatBreadth` <- as.factor(panth$`HabitatBreadth`)
panth$`Terrestriality` <- as.factor(panth$`Terrestriality`)
panth$`TrophicLevel` <- as.factor(panth$`TrophicLevel`)

panth %<>% dplyr::select(!c(`WeaningHeadBodyLen_mm`, References, `AdultForearmLen_mm`, `AdultBodyMass_g_EXT`, `NeonateBodyMass_g_EXT`, `WeaningBodyMass_g_EXT`, MSW_Species, MSW_Genus, MSW_Family, `GR_MaxLat_dd`, `GR_MinLong_dd`, `WeaningBodyMass_g`)) #too many NAs or just taxon
```


```{r categorize variables}


```


```{r make taxa dummy variables}
dummy_panth <- dummyVars(~ MSW_Order, data = panth)
panth_updated <- as_tibble(predict(dummy_panth, newdata = panth))
panth <- cbind(panth, panth_updated)
```


Measure PSR from gmpd

```{r gmpd psr}
gmpd_PSR <- gmpd %>% dplyr::select(HostCorrectedName,ParasiteCorrectedName) %>% 
  distinct() %>% 
  group_by(HostCorrectedName) %>% 
  summarize(psr=n())
panth %<>% mutate(HostCorrectedName = MSW_Binomial)
gmpd_panth <- gmpd_PSR %>% inner_join(panth, by="HostCorrectedName")
```

Prepare data by partitioning

```{r partition data}
set.seed(07981)
train <- sample(1:nrow(gmpd_panth), nrow(gmpd_panth) / 2)
test <- gmpd_panth %>% mutate(row_num = rownames(.)) %>% filter(!row_num %in% train) %>% dplyr::select(row_num)
test <- as.integer(test$row_num)
gmpd_panth_train <- gmpd_panth[train,]
gmpd_panth_test <- gmpd_panth[test,]
```

```{r gmpd train test datasets}
gmpd_panth_train <- as.data.frame(gmpd_panth_train)
gmpd_panth_test <- as.data.frame(gmpd_panth_test)
```


```{r gmpd initial model, warning = F}
brt1 <- gbm.step(data = gmpd_panth_train, gbm.x = 5:75, gbm.y= 2,
                 family = "poisson", tree.complexity = 5,
                 learning.rate = 0.004, bag.fraction = 0.5)
```

```{r}
summary(brt1)
```


```{r gmpd parasite type}
gmpd_macro <- gmpd %>% filter(ParType %in% c("Arthropod", "Helminth")) %>% dplyr::select(HostCorrectedName,ParasiteCorrectedName) %>% 
  distinct() %>% 
  group_by(HostCorrectedName) %>% 
  summarize(psr=n())

gmpd_micro <- gmpd %>% filter(ParType %in% c("Bacteria", "Protozoa", "Virus")) %>% dplyr::select(HostCorrectedName,ParasiteCorrectedName) %>% 
  distinct() %>% 
  group_by(HostCorrectedName) %>% 
  summarize(psr=n())

gmpd_panth_macro <- as.data.frame(gmpd_macro %>% inner_join(panth, by="HostCorrectedName"))
gmpd_panth_micro <- as.data.frame(gmpd_micro %>% inner_join(panth, by="HostCorrectedName"))
```

```{r macro micro model, warning=F}
brt2 <- gbm.step(data = gmpd_panth_macro, gbm.x = 5:75, gbm.y= 2,
                 family = "poisson", tree.complexity = 5,
                 learning.rate = 0.004, bag.fraction = 0.5)

brt3 <- gbm.step(data = gmpd_panth_micro, gbm.x = 5:75, gbm.y= 2,
                 family = "poisson", tree.complexity = 5,
                 learning.rate = 0.004, bag.fraction = 0.5)

```

```{r}
summary(brt2)
summary(brt3)
```
```{r gmpd pdp}
gbm.plot(brt1, n.plots=5, plot.layout=c(2, 3), write.title = FALSE)
#gbm.plot(brt2, n.plots=5, plot.layout=c(2, 3), write.title = FALSE)
#gbm.plot(brt3, n.plots=5, plot.layout=c(2, 3), write.title = FALSE)
```


```{r gmpd pdp2}
gbm.plot(brt1, variable.no = 17)
gbm.plot(brt2, variable.no = 17, plot.layout=c(2, 3), write.title = FALSE)
gbm.plot(brt3, variable.no = 17, plot.layout=c(2, 3), write.title = FALSE)
```

