---
title: "GMPD Pantheria"
author: "Daniel Suh"
date: "`r Sys.Date()`"
output: html_document
---

# Important note

Make sure to reclass any data using as.data.frame() before running gbm.step



```{r packages}
library(tidyverse)
library(magrittr)
library(here)

library(dismo)
library(caret)
library(gbm)

#source(here("previous work", "elith_tutorial", "brt.functions.R"))
```

Read in data. To start, we are focusing on gmpd and pantheria

```{r read}
gmpd <- read_csv(here("data/GMPD_main.csv"))
panth <- read_csv(here("data/panth_subset.csv"))
```


```{r make taxa dummy variables}
dummy_gmpd <- dummyVars(~ HostOrder, data = gmpd)
gmpd_updated <- as_tibble(predict(dummy_gmpd, newdata = gmpd))
gmpd <- cbind(gmpd, gmpd_updated)
gmpd_dummy_vars <- gmpd %>% dplyr::select(HostCorrectedName, HostOrderArtiodactyla, HostOrderCarnivora, HostOrderPerissodactyla, HostOrderPrimates) %>% distinct()
```


Measure PSR from gmpd

```{r gmpd psr}
gmpd_PSR <- gmpd %>% filter(ParType %in% c("Helminth", "Virus")) %>% dplyr::select(HostCorrectedName,ParasiteCorrectedName) %>% 
  distinct() %>% 
  group_by(HostCorrectedName) %>% 
  summarize(psr=n())
gmpd_PSR %<>% inner_join(., gmpd_dummy_vars)
panth %<>% mutate(HostCorrectedName = MSW_Binomial)
gmpd_panth <- gmpd_PSR %>% inner_join(panth, by="HostCorrectedName")
gmpd_panth %<>% dplyr::select(-c(HostCorrectedName, MSW_Order, MSW_Binomial))
```

Prepare data by partitioning

```{r partition data}
set.seed(07981)
train <- sample(1:nrow(gmpd_panth), nrow(gmpd_panth)*0.8)
test <- gmpd_panth %>% mutate(row_num = rownames(.)) %>% filter(!row_num %in% train) %>% dplyr::select(row_num)
test <- as.integer(test$row_num)
gmpd_panth_train <- gmpd_panth[train,]
gmpd_panth_test <- gmpd_panth[test,]
```

```{r gmpd train test datasets}
gmpd_panth_train <- as.data.frame(gmpd_panth_train)
gmpd_panth_test <- as.data.frame(gmpd_panth_test)
```


```{r gmpd initial model, warning = F}

brt1_elith_full <- gbm.step(data = gmpd_panth_train, gbm.x = 2:ncol(gmpd_panth_train), gbm.y= 1,
                 family = "poisson", tree.complexity = 5,
                 learning.rate = 0.004, bag.fraction = 0.5)
```

```{r}
summary(brt1_elith_full)
```


```{r helminth}
gmpd_helm <- gmpd %>% filter(ParType %in% c("Helminth")) %>% dplyr::select(HostCorrectedName,ParasiteCorrectedName) %>% 
  distinct() %>% 
  group_by(HostCorrectedName) %>% 
  summarize(psr=n())

gmpd_helm %<>% inner_join(., gmpd_dummy_vars)
```

```{r, arthropod}
gmpd_arth <- gmpd %>% filter(ParType %in% c("Arthropod")) %>% dplyr::select(HostCorrectedName,ParasiteCorrectedName) %>% 
  distinct() %>% 
  group_by(HostCorrectedName) %>% 
  summarize(psr=n())

gmpd_arth %<>% inner_join(., gmpd_dummy_vars)
```

```{r virus}
gmpd_virus <- gmpd %>% filter(ParType %in% c("Virus")) %>% dplyr::select(HostCorrectedName,ParasiteCorrectedName) %>% 
  distinct() %>% 
  group_by(HostCorrectedName) %>% 
  summarize(psr=n())



gmpd_virus %<>% inner_join(., gmpd_dummy_vars)
```

```{r bacteria}
gmpd_bact <- gmpd %>% filter(ParType %in% c("Bacteria")) %>% dplyr::select(HostCorrectedName,ParasiteCorrectedName) %>% 
  distinct() %>% 
  group_by(HostCorrectedName) %>% 
  summarize(psr=n())

gmpd_bact %<>% inner_join(., gmpd_dummy_vars)
```

```{r protozoa}
gmpd_proto <- gmpd %>% filter(ParType %in% c("Protozoa")) %>% dplyr::select(HostCorrectedName,ParasiteCorrectedName) %>% 
  distinct() %>% 
  group_by(HostCorrectedName) %>% 
  summarize(psr=n())

gmpd_proto %<>% inner_join(., gmpd_dummy_vars)
```


```{r gmpd virus}
gmpd_panth_helm <- as.data.frame(gmpd_helm %>% inner_join(panth, by="HostCorrectedName"))
gmpd_panth_virus <- as.data.frame(gmpd_virus %>% inner_join(panth, by="HostCorrectedName"))
gmpd_panth_arth <- as.data.frame(gmpd_arth %>% inner_join(panth, by="HostCorrectedName"))
gmpd_panth_bact <- as.data.frame(gmpd_bact %>% inner_join(panth, by="HostCorrectedName"))
gmpd_panth_proto <- as.data.frame(gmpd_proto %>% inner_join(panth, by="HostCorrectedName"))

```

Use only hosts that overlap between gmpd_panth_macro and gmpd_panth_micro
```{r}
gmpd_panth_helm_hosts <- gmpd_panth_helm$MSW_Binomial #get all host species from macro database
gmpd_panth_virus_hosts <- gmpd_panth_virus$MSW_Binomial #get remaining host species from micro database
gmpd_panth_arth_hosts <- gmpd_panth_arth$MSW_Binomial
gmpd_panth_bact_hosts <- gmpd_panth_bact$MSW_Binomial
gmpd_panth_proto_hosts <- gmpd_panth_proto$MSW_Binomial

intersect_hosts <- Reduce(intersect, list(gmpd_panth_helm_hosts, gmpd_panth_virus_hosts, gmpd_panth_arth_hosts, gmpd_panth_bact_hosts, gmpd_panth_proto_hosts))

gmpd_panth_helm %<>% filter(MSW_Binomial %in% intersect_hosts)
gmpd_panth_virus %<>% filter(MSW_Binomial %in% intersect_hosts)
gmpd_panth_arth %<>% filter(MSW_Binomial %in% intersect_hosts)
gmpd_panth_bact %<>% filter(MSW_Binomial %in% intersect_hosts)
gmpd_panth_proto %<>% filter(MSW_Binomial %in% intersect_hosts)

```



```{r}
gmpd_panth_helm %<>% dplyr::select(-c(HostCorrectedName, MSW_Order, MSW_Binomial))

train <- sample(1:nrow(gmpd_panth_helm), nrow(gmpd_panth_helm) * 0.8)
test <- gmpd_panth_helm %>% mutate(row_num = rownames(.)) %>% filter(!row_num %in% train) %>% dplyr::select(row_num)
test <- as.integer(test$row_num)
gmpd_panth_helm_train <- gmpd_panth_helm[train,]
gmpd_panth_helm_test <- gmpd_panth_helm[test,]
```

```{r}
gmpd_panth_arth %<>% dplyr::select(-c(HostCorrectedName, MSW_Order, MSW_Binomial))

train <- sample(1:nrow(gmpd_panth_arth), nrow(gmpd_panth_arth) * 0.8)
test <- gmpd_panth_arth %>% mutate(row_num = rownames(.)) %>% filter(!row_num %in% train) %>% dplyr::select(row_num)
test <- as.integer(test$row_num)
gmpd_panth_arth_train <- gmpd_panth_arth[train,]
gmpd_panth_arth_test <- gmpd_panth_arth[test,]
```

```{r}
gmpd_panth_virus %<>% dplyr::select(-c(HostCorrectedName, MSW_Order, MSW_Binomial))

train <- sample(1:nrow(gmpd_panth_virus), nrow(gmpd_panth_virus) * 0.8)
test <- gmpd_panth_virus %>% mutate(row_num = rownames(.)) %>% filter(!row_num %in% train) %>% dplyr::select(row_num)
test <- as.integer(test$row_num)

gmpd_panth_virus_train <- gmpd_panth_virus[train,]
gmpd_panth_virus_test <- gmpd_panth_virus[test,]
```

```{r}
gmpd_panth_bact %<>% dplyr::select(-c(HostCorrectedName, MSW_Order, MSW_Binomial))

train <- sample(1:nrow(gmpd_panth_bact), nrow(gmpd_panth_bact) * 0.8)
test <- gmpd_panth_bact %>% mutate(row_num = rownames(.)) %>% filter(!row_num %in% train) %>% dplyr::select(row_num)
test <- as.integer(test$row_num)
gmpd_panth_bact_train <- gmpd_panth_bact[train,]
gmpd_panth_bact_test <- gmpd_panth_bact[test,]
```

```{r}
gmpd_panth_proto %<>% dplyr::select(-c(HostCorrectedName, MSW_Order, MSW_Binomial))

train <- sample(1:nrow(gmpd_panth_proto), nrow(gmpd_panth_proto) * 0.8)
test <- gmpd_panth_proto %>% mutate(row_num = rownames(.)) %>% filter(!row_num %in% train) %>% dplyr::select(row_num)
test <- as.integer(test$row_num)
gmpd_panth_proto_train <- gmpd_panth_proto[train,]
gmpd_panth_proto_test <- gmpd_panth_proto[test,]
```

Elith method
```{r helminth model, warning=F}
# brt2_elith_macro <- gbm.step(data = gmpd_panth_macro_train, gbm.x = 2:47, gbm.y= 1,
#                  family = "poisson", tree.complexity = 5,
#                  learning.rate = 0.001, bag.fraction = 0.5)
# 
# brt3_elith_micro <- gbm.step(data = gmpd_panth_micro_train, gbm.x = 2:47, gbm.y= 1,
#                  family = "poisson", tree.complexity = 5,
#                  learning.rate = 0.001, bag.fraction = 0.5)

brt2_elith_helm <- gbm.step(data = gmpd_panth_helm_train, gbm.x = 2:ncol(gmpd_panth_train), gbm.y= 1,
                 family = "poisson", tree.complexity = 5,
                 learning.rate = 0.004, bag.fraction = 0.5)
```


```{r virus model, warning=F}
brt3_elith_virus <- gbm.step(data = gmpd_panth_virus_train, gbm.x = 2:ncol(gmpd_panth_train), gbm.y= 1,
                 family = "poisson", tree.complexity = 5,
                 learning.rate = 0.004, bag.fraction = 0.5)
```


```{r arthropod model, warning=F}
brt4_elith_arth <- gbm.step(data = gmpd_panth_arth_train, gbm.x = 2:ncol(gmpd_panth_train), gbm.y= 1,
                 family = "poisson", tree.complexity = 5,
                 learning.rate = 0.002, bag.fraction = 0.5)
```


```{r bacteria model, warning=F}
brt5_elith_bact <- gbm.step(data = gmpd_panth_bact_train, gbm.x = 2:ncol(gmpd_panth_train), gbm.y= 1,
                 family = "poisson", tree.complexity = 5,
                 learning.rate = 0.004, bag.fraction = 0.5)
```


```{r protozoa model, warning=F}
brt6_elith_proto <- gbm.step(data = gmpd_panth_proto_train, gbm.x = 2:ncol(gmpd_panth_train), gbm.y= 1,
                 family = "poisson", tree.complexity = 5,
                 learning.rate = 0.004, bag.fraction = 0.5)
```

```{r}
summary(brt1_elith_full)
summary(brt2_elith_helm)
summary(brt3_elith_virus)
summary(brt4_elith_arth)
summary(brt5_elith_bact)
summary(brt6_elith_proto)
```
```{r gmpd pdp 6}
gbm.plot(brt1_elith_full, n.plots=6, plot.layout=c(2, 3), write.title = FALSE, common.scale = T, main = "full")
gbm.plot(brt2_elith_helm, n.plots=6, plot.layout=c(2, 3), write.title = FALSE, common.scale = T, main = "helminth")
gbm.plot(brt3_elith_virus, n.plots=6, plot.layout=c(2, 3), write.title = FALSE, common.scale = T, main = "virus")
gbm.plot(brt4_elith_arth, n.plots=6, plot.layout=c(2, 3), write.title = FALSE, common.scale = T, main = "arthropod")
gbm.plot(brt5_elith_bact, n.plots=6, plot.layout=c(2, 3), write.title = FALSE, common.scale = T, main = "bacteria")
gbm.plot(brt6_elith_proto, n.plots=6, plot.layout=c(2, 3), write.title = FALSE, common.scale = T, main = "protozoa")
```



```{r gmpd pdp2}
brt1_names <- brt1_elith_full$var.names
brt2_names <- brt2_elith_helm$var.names
brt3_names <- brt3_elith_virus$var.names
brt4_names <- brt4_elith_arth$var.names
brt5_names <- brt5_elith_bact$var.names
brt6_names <- brt6_elith_proto$var.names

brt1_long_plot_no <- which(brt1_names %in% c("MaxLongevity_m"))
brt2_long_plot_no <- which(brt2_names %in% c("MaxLongevity_m"))
brt3_long_plot_no <- which(brt3_names %in% c("MaxLongevity_m"))
brt4_long_plot_no <- which(brt4_names %in% c("MaxLongevity_m"))
brt5_long_plot_no <- which(brt5_names %in% c("MaxLongevity_m"))
brt6_long_plot_no <- which(brt6_names %in% c("MaxLongevity_m"))

gbm.plot(brt1_elith_full, variable.no = brt1_long_plot_no, plot.layout=c(1, 1), write.title = FALSE, main = "full")
gbm.plot(brt2_elith_helm, variable.no = brt2_long_plot_no, plot.layout=c(1, 1), write.title = FALSE, main = "helm")
gbm.plot(brt3_elith_virus, variable.no = brt3_long_plot_no, plot.layout=c(1, 1), write.title = FALSE, main = "virus")
gbm.plot(brt4_elith_arth, variable.no = brt4_long_plot_no, plot.layout=c(1, 1), write.title = FALSE, main = "arth")
gbm.plot(brt5_elith_bact, variable.no = brt5_long_plot_no, plot.layout=c(1, 1), write.title = FALSE, main = "bact")
gbm.plot(brt6_elith_proto, variable.no = brt6_long_plot_no, plot.layout=c(1, 1), write.title = FALSE, main = "proto")
```

Becker and Han Method
```{r}
brt1_other_full <- gbm(psr ~ ., data = gmpd_panth_train,
                 distribution = "poisson", n.trees = brt1_elith_full$n.trees,
                 interaction.depth = 5, shrinkage = 0.004)

brt2_other_helm <- gbm(psr ~ ., data = gmpd_panth_helm_train,
                 distribution = "poisson", n.trees = brt2_elith_helm$n.trees,
                 interaction.depth = 5, shrinkage = 0.001)

brt3_other_virus <- gbm(psr ~ ., data = gmpd_panth_virus_train,
                 distribution = "poisson", n.trees = brt3_elith_virus$n.trees,
                 interaction.depth = 5, shrinkage = 0.001)
```

```{r}
summary(brt1_other_full)
summary(brt2_other_helm)
summary(brt3_other_virus)
```


```{r}
get_mse <- function(model, test) {
  pred <- predict(model, newdata = test, n.trees = model$n.trees)
  MSE <- mean((pred - test$psr)^2)
  RMSD <- sqrt(mean((pred-test$psr)^2))
  name <- print(substitute(model))
  print(paste("Mean Squared Error for", name,"= ", MSE))
  print(paste("Root Mean Square Deviation for", name, " = ", RMSD))
}
```


```{r}
get_mse(brt1_elith_full, gmpd_panth_test)
get_mse(brt2_elith_helm, gmpd_panth_helm_test)
get_mse(brt3_elith_virus, gmpd_panth_virus_test)
get_mse(brt4_elith_arth, gmpd_panth_arth_test)
get_mse(brt5_elith_bact, gmpd_panth_bact_test)
get_mse(brt6_elith_proto, gmpd_panth_proto_test)
```



Prediction. Mean Squared Error and Root Mean Square Deviation
```{r}
brt1_full_test_pred <- predict(brt1_elith_full, newdata = gmpd_panth_test, n.trees = brt1_elith_full$n.trees)
mean((brt1_full_test_pred-gmpd_panth_test$psr)^2)
sqrt(mean((brt1_full_test_pred-gmpd_panth_test$psr)^2))
```

```{r}
brt1_full_test_pred_other <- predict(brt1_other_full, newdata = gmpd_panth_test, n.trees = brt1_elith_full$n.trees)
mean((brt1_full_test_pred_other-gmpd_panth_test$psr)^2)
sqrt(mean((brt1_full_test_pred_other-gmpd_panth_test$psr)^2))
```

```{r}
brt2_helm_test_pred <- predict(brt2_elith_helm, newdata = gmpd_panth_helm_test, n.trees = brt2_elith_helm$n.trees)
mean((brt2_helm_test_pred-gmpd_panth_helm_test$psr)^2)
sqrt(mean((brt2_helm_test_pred-gmpd_panth_helm_test$psr)^2))
```

```{r}
# brt2_macro_test_pred_other <- predict(brt2_other_macro, newdata = gmpd_panth_macro_test, n.trees = brt2_other_macro$n.trees)
# mean((brt2_macro_test_pred_other-gmpd_panth_macro_test$psr)^2)
# sqrt(mean((brt2_macro_test_pred_other-gmpd_panth_macro_test$psr)^2))
```

```{r}
brt3_virus_test_pred <- predict(brt3_elith_virus, newdata = gmpd_panth_virus_test, n.trees = brt3_elith_virus$n.trees)
mean((brt3_virus_test_pred-gmpd_panth_virus_test$psr)^2)
sqrt(mean((brt3_virus_test_pred-gmpd_panth_virus_test$psr)^2))
```

```{r}
# brt3_micro_test_pred_other <- predict(brt3_other_micro, newdata = gmpd_panth_micro_test, n.trees = brt3_other_micro$n.trees)
# mean((brt3_micro_test_pred_other-gmpd_panth_micro_test$psr)^2)
# sqrt(mean((brt3_micro_test_pred_other-gmpd_panth_micro_test$psr)^2))
```
