---
title: "Pantheria PCA"
author: "Daniel Suh"
date: "`r Sys.Date()`"
output: html_document
---


Pantheria includes a wide variety of life history traits. In our analysis, we are most interested in those traits that map onto the fast-slow pace of life continuum. Longevity is one trait that maps well to pace of life but does not fully capture it. PCA might be worthwhile to reduce the dimensions of the data with the hope that a principal component that captures pace of life arises. We can then use that component in a regression model to see if pace of life correlates with PSR. 

To start, we can just look at the PC1 from Healy's analysis which is supposed to be capturing pace of life of organisms. Does this have a significant relationship with parasite species richness?

```{r, message = F}
library(tidyverse)
library(magrittr)
library(here)
```


```{r}
healy_axis <- read_csv(here("data/healy_axis_analysis.csv"))
healy_host <- read_csv(here("data/healy_host_par.csv"))
```

```{r}
healy_psr <- healy_host %>% dplyr::select(host, parasite) %>% group_by(host) %>% summarize(psr=n())
```

```{r}
healy_axis %<>% mutate(host = gsub("_", " ", healy_axis$species))
healy_axis_psr <- left_join(healy_axis, healy_psr, by = "host")

#average over duplicates
healy_axis_psr_avg <- healy_axis_psr %>% group_by(species) %>% mutate_each(funs(mean), c(2:15,17,18,20,24)) %>% distinct()

healy_axis_psr_avg$mode_of_life <- factor(healy_axis_psr_avg$mode_of_life)
healy_axis_psr_avg$IUCN <- factor(healy_axis_psr_avg$IUCN)
healy_axis_psr_avg$met_type <- factor(healy_axis_psr_avg$met_type)

healy_axis_psr_avg %<>% dplyr::select(-c(taxa_name,animal,host))
```

